---
title: "Mini-Project 02 - How Do You Do 'You Do You'?"
author: "anapau-tapia"
editor:
    mode: source
format:
    html:
        code-fold: true
---


## Introduction

How do Americans spend their time? 
Using data from the American Time Use Survey (ATUS), we will analyze, in this mini-project, the activities people across the U.S. like to do.

Here I intent to demonstrate my skills in: 

*Data Synthesis*: Combining disparate ATUS tables into a unified dataset.
*Visual Insights*: Highlighting trends through publication-quality plots and tables.
*Technical Rigor*: Transparent R code and documentation of the analytical process.


## Data Acquisition and Preparation

```{r}
library(tidyverse)
library(glue)
library(readxl)
library(httr2)

load_atus_data <- function(file_base = c("resp", "rost", "act")){
  if(!dir.exists(file.path("data", "mp02"))){
    dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
  }
  
  file_base <- match.arg(file_base)
  
  file_name_out <- file.path("data", 
                             "mp02", 
                             glue("atus{file_base}_0324.dat"))
  
  if(!file.exists(file_name_out)){
    
    url_end <- glue("atus{file_base}-0324.zip")
    
    temp_zip <- tempfile(fileext=".zip")
    temp_dir <- tools::file_path_sans_ext(temp_zip)
    
    request("https://www.bls.gov") |>
      req_url_path("tus", "datafiles", url_end) |>
      req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:143.0) Gecko/20100101 Firefox/143.0") |>
      req_perform(temp_zip)
    
    unzip(temp_zip, exdir = temp_dir)
    
    temp_file <- file.path(temp_dir, basename(file_name_out))
    
    file.copy(temp_file, file_name_out)
  }
  
       file <- read_csv(file_name_out,
                     show_col_types = FALSE,
                     n_max=50000)

  
  res <- switch(file_base,
         resp = file |> 
           rename(survey_weight = TUFNWGTP, 
                  time_alone = TRTALONE, 
                  survey_year = TUYEAR,
                  month = TUMONTH,
                  occ_code = TEIO1OCD,
                  hefaminc = any_of(c("HEFAMINC", "hefaminc"))) |>
           mutate(
            
             has_children = case_when(
               TRCHILDNUM > 0 ~ "Yes",
               TRUE ~ "No"
             ),
             
             marital_status = case_match(TRSPPRES,
                                         1 ~ "Married",
                                         2 ~ "Partnered",
                                         3 ~ "Single",
                                         .default = "Unknown"),
            
             employment_status = case_match(TELFS,
                                            1 ~ "Employed",
                                            2 ~ "Employed", # Employed-absent
                                            3 ~ "Unemployed",
                                            4 ~ "Unemployed", # Layoff
                                            5 ~ "Not in Labor force",
                                            .default = "Unknown"), 
             
             is_college_student = case_match(TESCHFT, 1 ~ TRUE, 2 ~ FALSE, .default = FALSE),
            
             is_retired = (TELFS == 5)
           ),
         rost = file |> 
           rename(age = any_of(c("TEAGE", "teage"))) |>
           mutate(sex = case_match(TESEX, 1 ~ "M", 2 ~ "F")),
         act = file |> 
           mutate(activity_n_of_day = TUACTIVITY_N, 
                  time_spent_min = TUACTDUR24, 
                  start_time = TUSTARTTIM, 
                  stop_time = TUSTOPTIME, 
                  level1_code = paste0(TRTIER1P, "0000"), 
                  level2_code = paste0(TRTIER2P, "00"), 
                  level3_code = TRCODEP,
                  # 7. Location of Activity 
                  location = case_match(TEWHERE,
                                        1 ~ "Home",
                                        2 ~ "Work",
                                        3 ~ "Someone else's home",
                                        4 ~ "Restaurant/Bar",
                                        5 ~ "Place of worship",
                                        6 ~ "Grocery store",
                                        12 ~ "Public Transit",
                                        .default = "Other/Travel"))
  ) 
  res|> 
    rename(participant_id = any_of(c("TUCASEID", "tucaseid"))) |>
    rename_with(tolower) |> 
    select(matches("[:lower:]", ignore.case = FALSE))
}

load_atus_activities <- function(){
  if(!dir.exists(file.path("data", "mp02"))){
    dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
  }
  
  dest_file <- file.path("data",
                         "mp02",
                         "atus_activity_codes.csv")
  
  if(!file.exists(dest_file)){
    download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                  quiet=TRUE, 
                  destfile=dest_file, 
                  mode="wb")
  }
  
  read_csv(dest_file, show_col_types = FALSE)
  
}
resp <- load_atus_data("resp")
act <- load_atus_data("act")
rost <- load_atus_data("rost")

activities <- load_atus_activities()


act |>
  inner_join(activities, join_by(level3_code == code_level3)) |> 
  filter(str_detect(task_level3, "Sleeping")) |> 
  summarize(total_hours = sum(time_spent_min) / 60)


resp |>
  inner_join(rost, join_by(participant_id == participant_id))

resp |>
  inner_join(rost, join_by(participant_id == participant_id)) |>
  filter(sex == "F", 
         survey_year == 2003, 
         time_alone == 0) |>
  summarize(total_count = n())

#3.1 Children under 18 living in respondent's home
resp |> 
  count(has_children)

#3.2 respondent's Marital status
resp |> 
  count(marital_status)

#3.3 Respondent’s employment status
resp |>
  group_by(employment_status) |>
  summarise(total_respondents = n())

#3.4 Location of activity
act |>
  count(location) |>
  arrange(desc(n)) |>
  
  head(5)

#3.5 Respondent's enrolled in University
resp |>
  summarise(
    college_student_count = sum(is_college_student, na.rm = TRUE),
    percent_college = mean(is_college_student, na.rm = TRUE) * 100
  )
```
### 4.1: Respondents answer ATUS #
```{r}
total_respondents <- resp |> 
  distinct(participant_id) |> 
  nrow()
```
since 2003, there have been `r resp|> distinct(participant_id) |> nrow()` different respondents who have answered the ATUS survey.


### 4.2: Sports Watching #
```{r}
sport_count <- activities |>
  filter(str_starts(code_level3, "1301")) |> 
  distinct(code_level3) |>
  nrow()
```
ATUS asked a total of `r sport_count` sports count different sport to american people to know which one they best enjoy to watch.


### 4.3: % Retired (Weighted) #
```{r}
retirement_stats <- resp |>
  summarize(
    percent_retired = weighted.mean(is_retired, w = survey_weight, na.rm = TRUE) * 100
  )
```  
Approximately `r round(retirement_stats$percent_retired, 2)`% of Americans are retired.


### 4.4: Average Sleep #
```{r}
sleep_per_person <- act |>
    filter(level2_code == "010100") |>
    group_by(participant_id) |>
    summarize(total_sleep_min = sum(time_spent_min, na.rm = TRUE))


sleep_stats <- resp |>
    left_join(sleep_per_person, by = "participant_id") |>
    mutate(total_sleep_min = replace_na(total_sleep_min, 0)) |>
    summarize(
        avg_sleep_hours = weighted.mean(total_sleep_min, w = survey_weight) / 60
    )
```  
On average, Americans spend `r round(sleep_stats$avg_sleep_hours, 2)` hours sleeping per night. 


### 4.5: Child Care #
```{r}

childcare_totals <- act |>
    filter(str_starts(level2_code, "0301|0302|0303")) |>
    group_by(participant_id) |>
    summarize(total_childcare_min = sum(time_spent_min, na.rm = TRUE))

childcare_results <- resp |>
    filter(has_children == "Yes") |> 
    left_join(childcare_totals, by = "participant_id") |>
    mutate(total_childcare_min = replace_na(total_childcare_min, 0)) |>
    summarize(
        avg_childcare_hours = weighted.mean(total_childcare_min, w = survey_weight) / 60
    )
```
For Americans with at least one child in the household, the average time spent on childcare activities is `r round(childcare_results$avg_childcare_hours, 2)` hours per day

### 5.1: lawn care #
```{r}
library(gt)

df_joined <- act |>
  left_join(resp, by = "participant_id") |>
  left_join(activities, by = c("level3_code" = "code_level3"))

df_joined |>
  filter(level3_code == "020102") |> 
  group_by(employment_status) |>
  filter(employment_status %in% c("Employed", "Not in Labor force")) |>
  summarize(avg_min = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE)) |>
  gt() |>
  tab_header(title = "Average Daily Minutes on Lawn Care 🌿",subtitle = "Comparison by Employment Status") |>
  fmt_number(columns = avg_min, decimals = 1) |>
  cols_label(employment_status = "Status", avg_min = "Avg. Minutes") |>
   tab_source_note(md(
      "According to the results from the ATUS survey, individuals who are not in the labor force spend 5.34% more time on lawn care than those who are employed. This suggests that unemployed individuals may utilize their additional free time to mow their lawns and tend to their gardens. "))

```
### 5.2 social/recreational activities #
```{r}
df_joined |>
  filter(str_starts(level3_code, "12")) |> # Socializing, Relaxing, and Leisure
  group_by(marital_status) |>
  summarize(avg_leisure_hrs = weighted.mean(time_spent_min, survey_weight) / 60) |>
  gt() |>
  tab_header(title = "Time Spent on Social/Recreational activities 🧩 ") |>
  fmt_number(columns = avg_leisure_hrs, decimals = 2) |>
  tab_source_note(md(
      "This table illustrates the amount of time people spend on social and recreational activities based on their marital status. According to the data, partnered individuals spend the most time on these activities, followed by single people. Married individuals spend the least amount of time on social and recreational activities compared to the other groups. "))

```


### 5.4 Activities on transit #
```{r}
df_joined |>
  filter(location == "Public Transit") |> 
  group_by(task_level3) |> 
  summarize(total_hours = sum(time_spent_min * survey_weight, na.rm = TRUE) / 60) |>
  slice_max(total_hours, n = 5) |>
  gt() |>
  tab_header(title = "Top 5 Activities on Public Transit 🚉 ") |>
  fmt_number(columns = total_hours, decimals = 0) |>
  cols_label(task_level3 = "Activity", total_hours = "Estimated Population Hours") |>
  tab_source_note(md(
      "With the data obtained from the ATUS survey, people declared that they mainly use the public transportation to go to work "))
```

###5.5 


###6.1 Household activities and age
```{r}
hh_age_data <- act |>
    filter(level1_code == "020000") |>
    # Use tulineno (lowercase) to identify the respondent
    inner_join(rost |> filter(tulineno == 1), by = "participant_id") |>
    inner_join(resp, by = "participant_id") |>
    group_by(age) |>
    summarize(avg_hh_time = weighted.mean(time_spent_min, survey_weight))

ggplot(hh_age_data, aes(x = age, y = avg_hh_time)) +
    geom_line(color = "steelblue") +
    geom_smooth(method = "loess", color = "red", se = FALSE) +
    labs(title = "Time Spent on Household Activities by Age",
         x = "Age", y = "Average Minutes per Day") +
    theme_minimal()

```
####
In this graphic, we analyzed how much time Americans spend on household activities. According to the results, teenagers (under age 20) and people in their 50s and 60s spend less time on these tasks than other age groups. Interestingly, the end of the graph shows that people over 70 see a positive increase in time spent, making them the most active group in this category..####

### 6.2 Summer Education Employees #
```{r}
df_joined |>
  # Codes 2200-2340 generally cover Education, Training, and Library Occupations
  filter(occ_code >= 2200 & occ_code <= 2340) |>
  filter(month %in% c(6, 7, 8)) |> # June, July, August
  group_by(task_level2) |>
  summarize(avg_min = weighted.mean(time_spent_min, survey_weight, na.rm = TRUE)) |>
  slice_max(avg_min, n = 10) |>
  ggplot(aes(x = reorder(task_level2, avg_min), y = avg_min, fill = avg_min)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "skyblue", high = "darkblue") +
  labs(title = "Summer Activities for Education Professionals ☀️",
       x = "Activity Category",
       y = "Avg Minutes/Day") +
  theme_minimal()

```
#### Education employees often enjoy the perk of having summers off, similar to students. From June to August, they have the opportunity to engage in their favorite leisure activities. This graph displays the ten most popular activities and the average minutes spent on each. Sleeping ranks as the number one activity, accounting for more than 225 minutes. Following that, the next three most popular activities—attending or hosting social events, attending sports, and working—occupy between 150 and 175 minutes. The remaining six activities are less common, with employees spending fewer than 150 minutes per day on them.####

###6.3 Education Activities 
```{R}
student_study <- act |>
    filter(level1_code == "060000") |>
    inner_join(resp, by = "participant_id") |>
    inner_join(rost |> filter(tulineno == 1), by = "participant_id") |>
    filter(is_college_student == TRUE) |>
    group_by(age) |>
    summarize(avg_study = weighted.mean(time_spent_min, survey_weight))

ggplot(student_study, aes(x = age, y = avg_study)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", color = "purple") +
    labs(title = "Study Time vs. Age for Enrolled Students",
         x = "Age", y = "Avg Minutes per Day") +
    theme_minimal()

```
#### Based on the sample of students interviewed for this survey, it appears that older students spend more time on educational activities such as studying, doing homework, and attending class. The data suggests a positive correlation: as age increases, so does the time spent on education. However, this trend may be influenced by outliers in the 30–35 age group, where some individuals reported over 400 minutes of daily educational activity ####

###6.4 Sport interests 
```{r}

library(ggbump)

bump_data <- act |>
    filter(level2_code == "130200") |>
    inner_join(rost |> filter(tulineno == 1), by = "participant_id") |>
    inner_join(resp, by = "participant_id") |>
    mutate(age_bin = cut(age, breaks = seq(15, 85, by = 10))) |>
    group_by(age_bin, level3_code) |>
    summarize(mean_time = weighted.mean(time_spent_min, survey_weight), .groups = 'drop') |>
    # --- ADD THE NAMES HERE ---
    mutate(sport_name = case_when(
        level3_code == "130203" ~ "Basketball",
        level3_code == "130205" ~ "Billiards",
        level3_code == "130206" ~ "Soccer",
        level3_code == "130207" ~ "Bowling",
        level3_code == "130213" ~ "Football",
        level3_code == "130214" ~ "Golfing",
        level3_code == "130216" ~ "Hockey",
        level3_code == "130220" ~ "Roller Skating",
        level3_code == "130221" ~ "Rugby",
        level3_code == "130226" ~ "Vehicle Touring",
        level3_code == "130227" ~ "Volleyball",
        level3_code == "130230" ~ "Weightlifting",
        level3_code == "130232" ~ "Wrestling",
        level3_code == "130299" ~ "Attending Sporting Events",
        TRUE ~ "Other Sport"
    )) |>
    group_by(age_bin) |>
    mutate(rank = rank(desc(mean_time), ties.method = "random")) |>
    filter(rank <= 5)

ggplot(bump_data, aes(x = age_bin, y = rank, color = sport_name, group = sport_name)) +
    geom_bump(linewidth = 1.5) +  
    geom_point(size = 3) +
    scale_y_reverse(breaks = 1:5) +
    labs(title = "Sports Popularity Rank by Age Group", 
         x = "Age Group", 
         y = "Rank",
         color = "Sport") + 
    theme_minimal()

```
####This bump chart shows that the sports people like to go watch really change as they get older. For younger people (ages 15 to 35), watching football is usually the #1 thing to do. But once people hit middle age, they start getting into all kinds of different sports instead of just the main ones. # 

###6.5 Time Spent per child
```{r}
Time_child_data <- act |>
    filter(level1_code == "030000") |>
    inner_join(resp, by = "participant_id") |>
    filter(trchildnum > 0) |>
    group_by(participant_id, trchildnum, survey_weight) |>
    summarize(total_care = sum(time_spent_min), .groups = 'drop') |>
    mutate(time_per_child = total_care / trchildnum) |>
    group_by(trchildnum) |>
    summarize(final_avg = weighted.mean(time_per_child, survey_weight))

ggplot(Time_child_data, aes(x = factor(trchildnum), y = final_avg)) +
    geom_col(fill = "coral") +
    labs(title = "Care Time Per Individual Child", x = "Number of Children", y = "Minutes per Child") +
    theme_minimal()

```
#### The histogram shows a clear negative trend. A single child receives the most one-on-one care time from their parents; however, as the number of children in a household increases, the care time per child decreases. As a result, a child in a family of seven receives less than 12 minutes of exclusive care, which is a huge contrast to the 80+ minutes spent on an only child ####   


## Final Insights and Deliverable
```{r}
library(gt)

respondent_info <- resp |>
    inner_join(rost |> 
                   filter(tulineno == 1) |> 
                   select(participant_id, age, sex), 
               by = "participant_id")

calc_market_stats <- function(df, label, target_offset = 30) {
    df |>
        group_by(participant_id, survey_weight) |>
        summarize(total_time = sum(time_spent_min), .groups = 'drop') |>
        summarize(
            Demographic = label,
            Low = quantile(total_time, 0.25),
            Typical = median(total_time),
            High = quantile(total_time, 0.75),
            App_Target = Typical + target_offset
        )
}

# Group 1: Stay-at-Home Dads
sahd_stats <- act |>
    filter(level1_code == "030000") |>
    inner_join(respondent_info |>
                   filter(sex == "M", trchildnum > 0, employment_status == "Not in Labor force"), 
               by = "participant_id") |>
    calc_market_stats("Stay-at-Home Dads")

# Group 2: Undergrads 
student_stats <- act |>
    filter(level1_code == "060000") |>
    inner_join(respondent_info |> filter(is_college_student == TRUE), by = "participant_id") |>
    calc_market_stats("College Undergraduates")

# Group 3: Remote Professionals
remote_stats <- act |>
    filter(level1_code == "050000", location == "Home") |>
    inner_join(respondent_info |> filter(employment_status == "Employed"), by = "participant_id") |>
    calc_market_stats("Remote Professionals")

# Final Table
bind_rows(sahd_stats, student_stats, remote_stats) |>
    gt() |>
    tab_header(title = "Market Research: Daily Time Use Targets 📟",
               subtitle = "Figures in minutes per day") |>
    fmt_number(columns = 2:5, decimals = 0) |>
  tab_source_note(md(
      "This table shows how much time three different groups of people—dads at home, college students, and remote workers—spend on their main daily tasks. We looked at the 'typical' amount of time they spend and then added 30 minutes to create an 'App Target.' This target is a goal for how much time we want people to spend using an app or doing an activity. It shows that you can't have the same goal for everyone, because a student's daily schedule looks totally different from a stay-at-home dad's schedule. "))

```


::: {.callout-note title="AI Usage Statement" collapse="true"}

Gemini was used to help write the code in this project, but all non-code text was generated without the use of any Generative AI tools. Additionally, Gemini was used to provide additional background information on the topic and to brainstorm ideas for the final open-ended prompt.

:::

------------------------------------------------------------------------

This work ©2025 by anapau-tapia was initially prepared as a Mini-Project for STA 9750 at Baruch College. More details about this course can be found at [the course site](https://michael-weylandt.com/STA9750) and instructions for this assignment can be found at [MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
